Binary Latent Representations for Efficient Ranking

### 2017-06-20 paper/paper.tex
Update with Movielens 1M results.

\begin{abstract}
Large-scale recommender systems often face severe latency and storage constraints at prediction time. These are particularly acute when the number of items that could be recommended is large, and calculating predictions for the full set is computationally intensive. To relax these constraints, we train recommendation models that use binary rather than real-valued user and item representations, and show that they offer a range of attractive speed-accuracy trade-offs. In particular, we demonstrate 13 times faster prediction at the expense of a modest 4\% loss in ranking accuracy on the Movielens 1M dataset.
\end{abstract}

@@

\section{Experiments}
\begin{table*}[htbp]
\begin{threeparttable}
\caption{Movielens 1M results}
\label{tb:results}
\centering
\begin{tabular}{rrrrrrrr}
\toprule
Dimension &   MRR & Binary MRR & MRR ratio\tnote{1} &   PPMS\tnote{2} & Binary PPMS & PPMS ratio & Memory use ratio \\
\midrule
       32 & 0.096 &      0.088 &     0.919 & 89,299 &     245,644 &      2.751 &            0.091 \\
       64 & 0.098 &      0.096 &     0.973 & 47,565 &     209,486 &      4.404 &            0.062 \\
      128 & 0.108 &      0.091 &     0.843 & 20,502 &     156,404 &      7.629 &            0.047 \\
      256 & 0.109 &      0.097 &     0.884 & 10,631 &     166,500 &     15.661 &            0.039 \\
      512 & 0.112 &      0.105 &     0.938 &   5391 &     110,418 &     20.479 &            0.035 \\
     1024 & 0.114 &      0.102 &     0.897 &   2689 &      69,349 &     25.786 &            0.033 \\
     2048 & 0.111 &      0.109 &     0.984 &   1352 &      34,826 &     25.745 &            0.032 \\
     4096 & 0.111 &      0.107 &     0.961 &    666 &      15,907 &     23.856 &            0.032 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small{
\item[1] Ratio of binary model MRR to real-valued model MRR
\item[2] Predictions Per Millisecond
}
\end{tablenotes}
\end{threeparttable}
\end{table*}
To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 1M dataset \citep{harper2016movielens}. The dataset contains 1 million ratings from 6000 users on 4000 movies. Because the computational speed improvements offered by the binary representations are linear in the size of the catalogue, the results on this dataset should be indicative of results that can be obtained on larger datasets. At the same time, the dataset's small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.

@@

As expected, for the same dimensionality a binary model achieves lower ranking accuracy. On average, the accuracy loss is around 8\%, varying between 2\% and 16\%.
Nevertheless, the model's speed enables some very favourable tradeoffs.

When optimizing for MRR in both settings (at 1024 dimensions for the continuous and 2048 for the binary case), the XNOR representation achieves 96\% of the maximum MRR of the float representations at 13 times the query throughput and using only 6\% of the memory.

When optimizing for query throughput at 32 dimensions for the continuous case, the XNOR representation (at 512 dimensions) can improve the ranking performance by achieving 110\% of the continuous MRR while being slightly faster at 1.2 times the query throughput, and using 55\% of the memory.

>>>>

\begin{abstract}
Large-scale recommender systems often face severe latency and storage constraints at prediction time. These are particularly acute when the number of items that could be recommended is large, and calculating predictions for the full set is computationally intensive. To relax these constraints, we train recommendation models that use binary rather than real-valued user and item representations, and show that they offer a range of attractive speed-accuracy trade-offs. In particular, we demonstrate 20 times faster prediction at the expense of a modest 7\% loss in ranking accuracy on the Movielens 1M dataset.
\end{abstract}

@@

\section{Experiments}
\begin{table*}[htbp]
\begin{threeparttable}
\caption{Movielens 1M results}
\label{tb:results}
\centering
\begin{tabular}{rrrrrrrr}
\toprule
Dimension &   MRR & Binary MRR & MRR ratio\tnote{1} &   PPMS\tnote{2} & Binary PPMS & PPMS ratio & Memory use ratio \\
\midrule
       32 & 0.072 &      0.059 &     0.816 & 85,510 &     264,310 &      3.091 &            0.091 \\
       64 & 0.075 &      0.065 &     0.862 & 44,964 &     221,000 &      4.915 &            0.062 \\
      128 & 0.076 &      0.070 &     0.926 & 22,394 &     166,273 &      7.425 &            0.047 \\
      256 & 0.078 &      0.072 &     0.916 & 12,040 &     189,353 &     15.727 &            0.039 \\
      512 & 0.079 &      0.073 &     0.921 &   5933 &     118,895 &     20.039 &            0.035 \\
     1024 & 0.081 &      0.076 &     0.933 &   2989 &      59,612 &     19.942 &            0.033 \\
     2048 & 0.080 &      0.075 &     0.930 &   1514 &      32,092 &     21.185 &            0.032 \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small{
\item[1] Ratio of binary model MRR to real-valued model MRR
\item[2] Predictions Per Millisecond
}
\end{tablenotes}
\end{threeparttable}
\end{table*}
To assess the accuracy-speed trade-offs enabled by the XNOR approach, we conduct an experiment on the Movielens 1M dataset \citep{harper2016movielens}. The dataset contains 1 million ratings from 6000 users on 4000 movies. Because the computational speed improvements offered by the binary representations are linear in the size of the catalogue, the results on this dataset should be indicative of results that can be obtained on larger datasets. At the same time, the dataset's small size enables us to easily test hyperparameter configurations and estimate models with high latent dimensions.

@@

As expected, for the same dimensionality a binary model achieves lower ranking accuracy. On average, the accuracy loss is around 10\%, varying between 7\% and 18\%.
Nevertheless, the binary model's speed enables some very favourable tradeoffs

When optimizing for MRR in both settings (at 1024 dimensions for the continuous and 1024 for the binary case), the XNOR representation achieves 93\% of the maximum MRR of the float representations at 20 times the query throughput and using only 3\% of the memory.

When optimizing for query throughput at 32 dimensions for the continuous case, the XNOR representation (at 512 dimensions) can improve the ranking performance by achieving 101\% of the continuous MRR while being slightly faster at 1.4 times the query throughput, and using 55\% of the memory.
